---
title: "Gapminder"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|-------------|------------------------------------------------------------------------------------------------------------------|----------------------------------------------------------------------------------------------------------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
glimpse(gapminder)
```

**Observations**:

-   Variable names: country, continent, year, LifeExp, Pop, gdpPercap

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`

year_max <- gapminder %>% 
  pull(year) %>% 
  max() 
year_min <- gapminder %>% 
  pull(year) %>% 
  min() 

year_max
year_min


```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year == year_min)  %>%
  select(continent, gdpPercap)   %>%
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_bar(stat = "identity") +
  ggtitle("Continent vs. GDP per Capita in 1952")

```

**Observations**:\

-   Write your observations here:
    -   It seems that out of the continents in the data set, Asia and Europe have the highest cumulative GDP per capita. Oceania has the lowest cumulative GDP per capita out of the continents in the dataset. This makes sense because Oceania has the least amount of countries among the continents with it only counting Australia and New Zealand.

**Difficulties & Approaches**:

-   The challenge with this bar graph is that we want to visualize GDP per capita per continent but larger continents have more countries and therefore larger GDP per capitas. I think that we could instead average the GDP per capita per country per continent if we wanted to compare each continent's GDP per capitas with one another.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2

gapminder %>%
  filter(year == year_min)  %>%
  select(continent, gdpPercap)   %>%
  ggplot(aes(x = continent, y = gdpPercap)) +
  scale_y_log10()+
  geom_boxplot() +
  ggtitle("Continent vs. GDP per Capita in 1952")

```

```{r}
gapminder %>%
  filter(year == year_min)  %>%
  group_by(continent)  %>%
  filter(gdpPercap >= (quantile(gdpPercap, 0.75) + 1.5*IQR(gdpPercap))) 
         
gapminder %>%
  filter(year == year_min)  %>%
  group_by(continent)  %>%
  filter(gdpPercap <= (quantile(gdpPercap, 0.25) - 1.5*IQR(gdpPercap))) 
```

**Observations**:

-   I found 9 outliers from the data set using the definition that outliers are countries with a gdp per capita greater than 1.5\*IQR of the third quartile and less than 1.5\*IQR of the first quartile within their continent grouping. I noticed that all the outliers within this data set are countries that have a much higher GDP per capita than the other countries in the continent grouping. Much of the outliers come from Africa and the Ameriacs, with two in Asia and one in Europe.

    ## Identify the outlier countries from q2

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  scale_y_log10() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2) 
  
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent

gapminder %>%
  filter(year == min(year)) %>%

  ggplot(aes(continent, gdpPercap)) +
  geom_boxplot() +
  scale_y_log10()+
  geom_point(
    data = . %>% filter(country %in% c("Angola", "Bahrain", "Canada", "Gabon", "Kuwait", "South Africa", "Switzerland", "United States", "Venezuela")),
    mapping = aes(color = country),
    size = 2
  ) + 
    ggtitle('Continent vs. GDP per capita in 1952') 
```

```{r}

gapminder %>%
  filter(year == year_max)  %>%
  group_by(continent)  %>%
  filter(gdpPercap >= (quantile(gdpPercap, 0.75) + 1.5*IQR(gdpPercap))) 
         
gapminder %>%
  filter(year == year_max)  %>%
  group_by(continent)  %>%
  filter(gdpPercap <= (quantile(gdpPercap, 0.25) - 1.5*IQR(gdpPercap))) 


```

**Observations**:

-   In the 1952 Continent vs. GDP per capita plot, the country that by far was the largest outlier within its continent was Kuwait within Asia.
-   In the 2007 Continent vs. GDP per capita plot, the outliers were predominantly found in Africa and Asia. The largest outlier within a continent would go to the United states within the Americas continent.

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph

gapminder %>%
  filter(continent == 'Oceania') %>%
  ggplot(aes(year, lifeExp, color = as_factor(country))) +
  geom_point(size = 2) +
  theme_minimal()
```

-   When looking at the Life expectancy of the continent of Oceania over time, we see a general upwards trend from 1950 to 2007. After 1970, Australia has had a slightly higher life expectancy than New Zealand.

```{r q5-task2}
## TASK: Your second graph

gapminder %>%
  filter(year == year_max) %>%
  ggplot(aes(lifeExp, gdpPercap)) +
  geom_point(size = 2) +
  ggtitle("Life Expectancy vs. GDP Per Capita in the year 2007")
```

-   It seems that when we plot life expectancy against gdp per capita for countries in 2007, there seem to be a general positive correlation after the country reaches a life expectancy of 70 or more. Before the life expectancy of 70, there seems to be little or no correlation between life expectancy and gdp per capita. From this chart we can assume that countries with lower gdp per capita have a wide range of life expectancies from 40 to 70.

```{r q5-task3}
## TASK: Your third graph

gapminder %>%
  filter(country == "Philippines") %>%
  ggplot(aes(year, gdpPercap))+
  geom_point(size = 2) +
  theme_minimal() + 
  ggtitle("Philippines GDP Per capita over time")
```

-   Philippines has been increasing in GDP per capita over time with an exception during the mid 1980's when the trend suffered a dip. According to Philippine history, the mid late 70s was a period of high economic growth due to the world's increased demand for commodities such as Philippine coconut and sugar. This was followed by a recession in the 80s due to a combination of lavish spending under the Marcos Regime, political instability from former president Ninoy's assassination, and a rise in global oil prices.
